---
title: "Bovid astragalus ecomorphology as a proxy for sub-member scale habitat variability
  in Shungura Formation member G"
author: "Madeleine Kelly"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Required Packages
```{r}
library(tidyverse)
library(geomorph)
library(Morpho)
library(MASS)
library(PCDimension) 
library(vegan)
library(patchwork)
```

### Load Scripts for Plotting Code
```{r}
source(file = "code/plotting_functions.R", local = TRUE) # this file includes homemade analysis and plotting functions
```

### Load Datasets
```{r}
# species data & other museum database information for modern astragali
modern_astragali_info <- read_csv(file = "data/modern_astragali_fmnh.csv")

# species data & other museum database information for fossil astragali
fossil_astragali_info <- read_csv(file = "data/complete_memberG_astragali.csv")

# 3D array of landmark coordinates from Checkpoint for modern (FMNH, n = 136) and fossil (NME, n = 195) astragali (total n = 331)
all_landmarks <- readRDS(file = "data/FMNH_NME_astragali_landmarks.rds")
```

Mean number of wild/adult modern individuals per species:

```{r}
species_counts <- modern_astragali_info %>% group_by(Genus_species) %>% tally()
mean(species_counts$n)
sd(species_counts$n)
```

### Generalized Procrustes Analysis
This is conducted on the entire (modern + fossil) dataset to ensure all data are comparable in their coordinates. 
```{r}
GPA_all <- gpagen(all_landmarks, PrinAxes = TRUE, ProcD = TRUE, Proj = TRUE)

summary(GPA_all)
plot(GPA_all)
```

Is there an effect of side?
```{r}
# Procrustes ANOVA on shape ~ side in all individuals:
summary(procD.lm(GPA_all$coords[,,1:136] ~ modern_astragali_info$Side))
```

### Principal Components Analysis

PCA using only modern astragali (n = 136). This produces PC axes that are independent of fossil data, so that when using PCs as predictors later (in CVA), the resulting predictions for the fossils are not influenced by the fossils themselves.

```{r principal components analysis of Procrustes coordinates, include=TRUE}
# PCA on modern data only (the first 136 specimens in the landmark dataset; also could be defined as anything of dimnames(all_landmarks)[[3]] that begins with "FMNH")
PCA_modern <- gm.prcomp(GPA_all$coords[,,1:136])
summary(PCA_modern)

# create matrix of PC scores
mod_PC_scores <- PCA_modern[["x"]]

# add PC scores to full dataframe of specimen information
modern_astragali_info <- bind_cols(modern_astragali_info, as.data.frame(mod_PC_scores))
```

#### Broken Stick: Which PCs explain more variation than expected by chance?
```{r}
# extract proportion of variance values for each principal component
pov <- as.data.frame(PCA_modern$sdev^2/sum(PCA_modern$sdev^2))
colnames(pov) <- "prop_variance"

# broken stick algorithm for determining what the proportion of variance is expected based on chance alone for each PC
pov$brokenStick <- brokenStick(1:nrow(pov),nrow(pov))
pov <- pov%>%
  mutate(keep = ifelse(brokenStick > prop_variance, 0, 1))%>%
  filter(keep == 1)
pov
```

#### Figure 2: PC1 vs. PC2 and PC1 vs. PC3

Bivariate plots of PC space, with modern data only:
```{r, echo = F, fig.width=9.5, fig.height=7}
#PC1 vs PC2
# convex hulls 
hull_12 <- modern_astragali_info %>% group_by(Habitat) %>%
  slice(chull(Comp1, Comp2))
#plot
pc1v2 <- ggplot()+ 
  geom_polygon(data = hull_12, aes(x = Comp1, y = Comp2, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ # plot convex hulls
  scale_fill_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open")) +
  geom_point(data = modern_astragali_info, mapping = aes(x = Comp1, y = Comp2, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  tribe_shapes() +
  theme_classic(base_size = 22) +
  xlab("PC1 (17.7%)")+
  ylab("PC2 (14.6%)")

#PC1 vs PC3
# convex hulls (PC1 vs PC3)
hull_13 <- modern_astragali_info %>% group_by(Habitat) %>%
  slice(chull(Comp1, Comp3))
#plot
pc1v3 <- ggplot()+ 
  geom_polygon(data = hull_13, aes(x = Comp1, y = Comp3, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ # plot convex hulls 
  scale_fill_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open")) +
  geom_point(data = modern_astragali_info, mapping = aes(x = Comp1, y = Comp3, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  tribe_shapes() +
  theme_classic(base_size = 22) +
  xlab("PC1 (17.7%)")+
  ylab("PC3 (7.2%)")

# show plots
pc1v2
pc1v3
```

3D, color-coded warps of minimum and maximum values along PC1, PC2, and PC3:
```{r}
# generate mesh of mean shape by warping specimen closest to the mean to the mean shape from the GPA
ref_spec <- mshape(GPA_all$coords[,,1:136]) # find mean shape of modern astragali
mean_spec <- findMeanSpec(GPA_all$coords[,,1:136]) # actual specimen closest to mean
mean_filename <- modern_astragali_info$filename[mean_spec] # filename of mean specimen
mean_filename <- gsub("\\..*",'', mean_filename) # removes the ".csv" from the end of the filename (removes everything after the period)

# read in .ply file of mean specimen
mean_mesh <- read.ply(file = paste("data/astragalus/ply/",mean_filename,".ply", sep = ""), ShowSpecimen = TRUE)

# non-Procrustes landmarks for actual specimen closest to the mean
mean_landmarks <- all_landmarks[,,mean_spec]

# warp real specimen closest to the mean to the mean shape
mean_warp <- warpRefMesh(mesh = mean_mesh, mesh.coord = mean_landmarks, ref = ref_spec, centered = F)
plot3d(mean_warp, aspect = F) 
```

On all color coded mesh warps, red shades are areas where either the max or min shape on the PC axis expands outward from the mean along that axis, while blues indicate areas where the shape contracts inward from the mean. All .ply files can be exported out of R and visualized in any 3D viewing software (e.g., Meshlab).
```{r}
# get minimum and max, extract from shapes list in the PCA output
PC1_min <- PCA_modern$shapes[[1]]$min
PC1_max <- PCA_modern$shapes[[1]]$max

PC2_min <- PCA_modern$shapes[[2]]$min
PC2_max <- PCA_modern$shapes[[2]]$max

PC3_min <- PCA_modern$shapes[[3]]$min
PC3_max <- PCA_modern$shapes[[3]]$max

# PC1 min warp 
warp_pc1_min <- plotRefToTarget(ref_spec, PC1_min, mesh = mean_warp, method = "surface")
meshDist(x = warp_pc1_min, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)

# PC1 max warp 
warp_pc1_max <- plotRefToTarget(ref_spec, PC1_max, mesh = mean_warp, method = "surface")
meshDist(x = warp_pc1_max, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)

# PC2 min warp
warp_pc2_min <- plotRefToTarget(ref_spec, PC2_min, mesh = mean_warp, method = "surface")
meshDist(x = warp_pc2_min, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)

# PC2 max warp
warp_pc2_max <- plotRefToTarget(ref_spec, PC2_max, mesh = mean_warp, method = "surface")
meshDist(x = warp_pc2_max, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)

# PC3 min warp
warp_pc3_min <- plotRefToTarget(ref_spec, PC3_min, mesh = mean_warp, method = "surface")
meshDist(x = warp_pc3_min, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)

# PC3 max warp
warp_pc3_max <- plotRefToTarget(ref_spec, PC3_max, mesh = mean_warp, method = "surface")
meshDist(x = warp_pc3_max, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)
```

### Allometry

Replicating some of the methods of Orgebin et al. (2025) to investigate correlation between centroid size and shape overall.

First, generate species mean shapes to investigate this:
```{r}
# list of unique species included in the analysis
species <- unique(modern_astragali_info$Genus_species) 

# blank array to store mean shapes
mean_shapes <- array(dim = c(20,3,length(species)))

# generate mean shapes
for (i in 1:length(species)) {
  who <- species[i] # species
  ind <- which(modern_astragali_info$Genus_species == who) # which rows correspond to the species in question
  ids <- modern_astragali_info[ind,]$filename
  spec_mean_i <- mshape(GPA_all$coords[,,ids]) # mean Procrustes aligned shape of species in question (dimnames in GPA_all match filename column in modern_astragali_info)
  mean_shapes[,,i] <- spec_mean_i # add mean shape to array
}
dimnames(mean_shapes)[[3]] <- species # name them correctly

# extract centroid size from GPA (all individuals)
modern_astragali_info$centroid_size <- GPA_all$Csize[1:136]

# species data (habitat and mean centroid size)
species_info <- modern_astragali_info %>% dplyr::select(Genus_species, Habitat, centroid_size) %>%
  group_by(Genus_species) %>% mutate(mean_centroid_size = mean(centroid_size))
species_info <- species_info %>% dplyr::select(Genus_species, Habitat, mean_centroid_size) %>% distinct()
# make sure these rows are in the same order as in GPA coords (they should be, but fix if needed)
```

##### Procrustes ANOVA (shape ~ centroid size):
```{r}
# regress species mean Procrustes shapes against centroid size (log-transformed)
summary(procD.lm(mean_shapes ~ log(species_info$mean_centroid_size)))
```

Log centroid size is significantly correlated with astragalus shape overall (but the R-squared value is relatively low, ~0.16). 

##### Variation Partitioning Analysis
Is size correlated with habitat? Orgebin et al. use a variation partitioning method to explore this (`varpart` function in vegan package).
```{r}
# based on the Orgebin et al. (2025) code, to run the variation partitioning analysis, the input data are a 2D array of the Procrustes coordinates, plus corresponding habitat and centroid size data. This is done on species means to account for uneven sampling

# varpart analysis 
gpa_2d_mod_means <- two.d.array(mean_shapes) #GPA coords of modern bovid species means
var_part_analysis_means <- varpart(gpa_2d_mod_means, ~species_info$Habitat, ~log(species_info$mean_centroid_size))
var_part_analysis_means
summary(var_part_analysis_means)
```
The variation partitioning analysis shows that habitat explains 4% of the shape variation, and log(centroid size) explains 12%, with habitat and log(centroid size) jointly explaining 1% of the shape variation in the sample. 

###### Supplementary Figure 1 (variation partitioning analysis):
```{r}
# plot Venn diagram from variation partitioning analysis (SOM Figure 1)
plot(var_part_analysis_means, Xnames=c("Habitat","Size"), cutoff = -Inf)
```

##### Kruskal-Wallis comparing centroid size among habitat groups

Does centroid size vary among habitat groups?
```{r}
# species mean centroid sizes
sp_mean_sizes <- modern_astragali_info %>% dplyr::select(Genus_species, Habitat, centroid_size) %>%
  group_by(Genus_species) %>% mutate(mean_centroid_size = mean(centroid_size))
sp_mean_sizes <- sp_mean_sizes %>% dplyr::select(Genus_species, Habitat, mean_centroid_size) %>% distinct()

# is there a significant difference among groups (Kruskal-Wallis test)
kruskal.test(x = log(sp_mean_sizes$mean_centroid_size), g = sp_mean_sizes$Habitat)
```

Examining log species mean centroid sizes (to control for uneven sampling across species), there is no significant difference among habitat groups (p = 0.44).

###### Figure 3: boxplots of centroid size among habitat groups
```{r}
# plot
ggplot(data = sp_mean_sizes) +
  geom_boxplot(aes(x = Habitat, y = log(mean_centroid_size), fill = Habitat, color = Habitat), alpha = 0.6)+
  scale_fill_brewer(palette = "BrBG", direction = -1)+
  scale_color_brewer(palette = "BrBG", direction = -1)+
  ylab("log(centroid size)")+
  theme_classic(base_size = 12)+
  theme(legend.position = "none")
ggsave("figures/Figure_2.png", width = 7, height = 7)
```

##### Regressing each PC against centroid size:
```{r relationship of PCs to log centroid size, include=TRUE}
# regress centroid size on all PCs and save R^2 values in dataframe
pc_vs_csize <- list()
pc_vs_csize_r2 <- data.frame(r2 = integer(), f_stat = integer(), pval_fstat = integer())
for (i in 1:53) {
  pc_n <- paste("Comp", i, sep = "")
  reg <- summary(lm(unlist(modern_astragali_info[,pc_n]) ~ log(modern_astragali_info$centroid_size)))
  pc_vs_csize[[i]] <- reg # save entire lm output in a list
  pc_vs_csize_r2[i,"r2"] <- reg$r.squared # extract R^2 value of regression
  pc_vs_csize_r2[i, "f_stat"] <- reg$fstatistic[1] # extract F-statistic
  pc_vs_csize_r2[i, "pval_fstat"] <- pf(reg$fstatistic[1], reg$fstatistic[2], reg$fstatistic[3], lower.tail = F) # calculate p-value of F-statistic
  rownames(pc_vs_csize_r2)[i] <- pc_n
}

# which PCs have a significant (p < 0.05) correlation with centroid size?
pc_vs_csize_r2$signif <- pc_vs_csize_r2$pval_fstat < 0.05

# plot PCs that are significantly correlated with centroid size
# PC1
pc1vCsize <- ggplot(data = modern_astragali_info, aes(x = Comp1, y = log(centroid_size)))+
  geom_point(aes(shape = Tribe, color = Habitat), size = 3)+
  tribe_shapes()+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_smooth(method = "lm", color = "gray40")+
  annotate("text", label = "R^2 == 0.50", x = 0.06, y= 3.5, parse = T)+
  labs(x = "PC1 (17.2%)", y = "log(Centroid Size)")+
  theme_classic(base_size = 12)

#PC3
pc3vCsize <- ggplot(data = modern_astragali_info, aes(x = Comp3, y = log(centroid_size)))+
  geom_point(aes(shape = Tribe, color = Habitat), size = 3)+
  tribe_shapes()+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_smooth(method = "lm", color = "gray40")+
  annotate("text", label = "R^2 == 0.067", x = 0.06, y= 3.5, parse = T)+
  labs(x = "PC3 (7.2%)", y = "log(Centroid Size)")+
  theme_classic(base_size = 12)

#PC10
pc10vCsize <- ggplot(data = modern_astragali_info, aes(x = Comp10, y = log(centroid_size)))+
  geom_point(aes(shape = Tribe, color = Habitat), size = 3)+
  tribe_shapes()+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_smooth(method = "lm", color = "gray40")+
  annotate("text", label = "R^2 == 0.035", x = 0.04, y= 3.5, parse = T)+
  labs(x = "PC10 (3.0%)", y = "log(Centroid Size)")+
  theme_classic(base_size = 12)

#PC12
pc12vCsize <- ggplot(data = modern_astragali_info, aes(x = Comp12, y = log(centroid_size)))+
  geom_point(aes(shape = Tribe, color = Habitat), size = 3)+
  tribe_shapes()+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_smooth(method = "lm", color = "gray40")+
  annotate("text", label = "R^2 == 0.044", x = -0.03, y= 3.5, parse = T)+
  labs(x = "PC12 (2.5%)", y = "log(Centroid Size)")+
  theme_classic(base_size = 12)

#PC17
pc17vCsize <- ggplot(data = modern_astragali_info, aes(x = Comp17, y = log(centroid_size)))+
  geom_point(aes(shape = Tribe, color = Habitat), size = 3)+
  tribe_shapes()+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_smooth(method = "lm", color = "gray40")+
  annotate("text", label = "R^2 == 0.065", x = 0.03, y= 3.5, parse = T)+
  labs(x = "PC17 (1.4%)", y = "log(Centroid Size)")+
  theme_classic(base_size = 12)
```


### Canonical Variates Analysis

First, using modern data only to test predictive ability on specimens where we know their habitat. PC scores along each individual axis are scaled between 0 and 1 to allow for clearer interpretation of loadings along CV axes.

```{r}
# vector of grouping variable (habitat categories)
cvagroups <- modern_astragali_info$Habitat

# scale each PC score column so that all values are between 0 and 1 -- this will not change relationships but will make the DFA loadings easier to interpret
min_max <- function(x){
  res <- (x - min(x))/(max(x)-min(x))
  return(res)
}
modern_PC_scores_scaled <- apply(mod_PC_scores, 2, min_max) # apply the scaling function to each column of the dataset
```


#### Iterative CVA

What is the influence of increasing the number of PCs one by one on cross-validation success:

```{r iterative CVA}
iter_cva <- list()
xvals <- list()
prop_correct <- data.frame(n_pc = integer(), loo_correct = integer())
for (i in 2:53) {
  cva_res <- lda(modern_PC_scores_scaled[,1:i], cvagroups, CV = TRUE)
  iter_cva[[i-1]] <- cva_res
  xvals[[i-1]] <- table(cvagroups, cva_res$class)
  prop_correct[i-1, "n_pc"] <- i
  prop_correct[i-1, "loo_correct"] <- sum(diag(prop.table(table(cvagroups, cva_res$class))))
}
```

##### Supplementary Figure 2: Iterative CVA

Visualise change in LOO cross-validation success across all CVAs:
```{r plot iterative CVA LOO CV results}
ggplot(data = prop_correct)+
  geom_point(aes(x = n_pc, y = loo_correct), color = alpha("black", alpha = 0.8))+
  geom_smooth(aes(x = n_pc, y = loo_correct), method = "loess", color = "mediumpurple4", fill = alpha("mediumpurple3", alpha = 0.6))+
  geom_hline(aes(yintercept = max(loo_correct)), linetype = 2)+ # maximum classification success
  labs(x = "Number of PCs", y = "Leave-one-out cross-validation success rate")+
  theme_classic()
```

#### CVA with all PCs:
```{r}
# conduct the CVA (without cross-validation) on scaled PC scores:
cvaresults_scaled <- lda(modern_PC_scores_scaled, cvagroups, CV=FALSE)
A_scaled <- cvaresults_scaled$scaling # loadings along each CV axis; used to interpret which PCs contribute the most to separation along each axis (Supp. Table 3)
grpmeans_scaled <- cvaresults_scaled$means

# calculate proportion of between-group variance explained by each CVA axis
proportion_of_trace <- cvaresults_scaled$svd^2 / sum(cvaresults_scaled$svd^2)
proportion_of_trace

# project modern specimens onto CVA
C_scaled <- modern_PC_scores_scaled%*%A_scaled 

# add specimen info to CV scores for plotting
C_scaled_info <- data.frame(CV1 = C_scaled[,1],
                     CV2 = C_scaled[,2],
                     CV3 = C_scaled[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe)
```

##### Leave-one-out cross-validation (Table 1):
```{r}
cvaresults_scaled_CV <- lda(modern_PC_scores_scaled, cvagroups, CV=TRUE)
crossval <- table(cvagroups, cvaresults_scaled_CV$class) # confusion matrix of specimen assignments
sum(diag(prop.table(crossval))) # total percent correct

# Table 1
altcrossval <- rbind(crossval[1,]/sum(crossval[1,]), crossval[2,]/sum(crossval[2,]), crossval[3,]/sum(crossval[3,]), crossval[4,]/sum(crossval[4,]))
dimnames(altcrossval) <- list("Actual"=c("F", "HC", "LC", "O"), "Predicted (CV)"=c("F", "HC", "LC", "O"))
print(round(altcrossval, 3))
```

Examine misclassification rates among tribes and species:
```{r}
cv_predictions <- as.data.frame(cvaresults_scaled_CV$posterior)
cv_predictions <- bind_cols(cv_predictions, modern_astragali_info[,c("Tribe", "Genus_species", "Habitat")])

# create a predicted habitat column to indicate the habitat that has the highest posterior probability of group membership for each astragalus
cv_predictions$pred_habitat <- colnames(cv_predictions)[apply(cv_predictions[,1:4], 1, which.max)]
cv_predictions$match <- (cv_predictions$Habitat == cv_predictions$pred_habitat)

# proportions of correct (TRUE) and incorrect (FALSE) classifications by tribe
tribe_prediction_performance <- table(cv_predictions$Tribe, cv_predictions$match)
tribe_prediction_performance_prop <- prop.table(tribe_prediction_performance, 1)

# proportions of correct (TRUE) and incorrect (FALSE) classifications by species
species_prediction_performance <- table(cv_predictions$Genus_species, cv_predictions$match)
species_prediction_performance_prop <- prop.table(species_prediction_performance, 1)
```

##### Figure 4: CV1 vs. CV2 and CV1 vs. CV3
```{r, echo = F, fig.width=9.5, fig.height=7}
# CV1 vs. CV2
hull_cva_12 <- C_scaled_info %>% group_by(Habitat) %>%
  slice(chull(CV1, CV2))
cva_1v2 <- ggplot()+
  geom_polygon(data = hull_cva_12, aes(x = CV1, y = CV2, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ # plot convex hulls
  scale_fill_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_point(data = C_scaled_info, mapping = aes(x = CV1, y = CV2, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  tribe_shapes() +
  theme_classic(base_size = 22) +
  xlab("CV1")+
  ylab("CV2")

# CV1 vs. CV3
hull_cva_13 <- C_scaled_info %>% group_by(Habitat) %>%
  slice(chull(CV1, CV3))
cva_1v3 <- ggplot()+
  geom_polygon(data = hull_cva_13, aes(x = CV1, y = CV3, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ # plot convex hulls
  scale_fill_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_point(data = C_scaled_info, mapping = aes(x = CV1, y = CV3, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  tribe_shapes() +
  theme_classic(base_size = 22) +
  xlab("CV1")+
  ylab("CV3")

# show plots
cva_1v2
cva_1v3
```


#### CVA with first 30 PCs:

The first 30 PCs capture 95% of the sample variance; this is a common cutoff used in geometric morphometric analyses.
```{r}
# conduct the CVA (without cross-validation):
cvaresults_30pc <- lda(modern_PC_scores_scaled[,1:30], cvagroups, CV=FALSE)
A_30pc <- cvaresults_30pc$scaling # Supplementary Table 5
grpmeans_30pc <- cvaresults_30pc$means

# project specimens onto CVA
C_30pc <- modern_PC_scores_scaled[,1:30]%*%A_30pc 

# add specimen info for plotting
C_30pc_info <- data.frame(CV1 = C_30pc[,1],
                     CV2 = C_30pc[,2],
                     CV3 = C_30pc[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe)
```

##### Leave-one-out cross-validation (Supplementary Table 6)

```{r}
cvaresults_30pc_CV <- lda(modern_PC_scores_scaled[,1:30], cvagroups, CV=TRUE)
crossval_30pc <- table(cvagroups, cvaresults_30pc_CV$class) # confusion matrix of specimen assignments
sum(diag(prop.table(crossval_30pc))) # total percent correct

# Supplementary Table 4
altcrossval_30pc <- rbind(crossval_30pc[1,]/sum(crossval_30pc[1,]), crossval_30pc[2,]/sum(crossval_30pc[2,]), crossval_30pc[3,]/sum(crossval_30pc[3,]), crossval_30pc[4,]/sum(crossval_30pc[4,]))
dimnames(altcrossval_30pc) <- list("Actual"=c("F", "HC", "LC", "O"), "Predicted (CV)"=c("F", "HC", "LC", "O"))
print(round(altcrossval_30pc, 3))
```

#### CVA with first 8 PCs (broken stick significant):

```{r}
# conduct the CVA (without cross-validation):
cvaresults_8pc <- lda(modern_PC_scores_scaled[,1:8], cvagroups, CV=FALSE)
A_8pc <- cvaresults_8pc$scaling # Supplementary Table 7
grpmeans_8pc <- cvaresults_8pc$means

# project specimens onto CVA
C_8pc <- modern_PC_scores_scaled[,1:8]%*%A_8pc #usually this would be M%*%A

# add specimen info for plotting
C_8pc_info <- data.frame(CV1 = C_8pc[,1],
                     CV2 = C_8pc[,2],
                     CV3 = C_8pc[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe)
```

##### Leave-one-out cross-validation (Supplementary Table 8)
```{r}
cvaresults_8pc_CV <- lda(modern_PC_scores_scaled[,1:8], cvagroups, CV=TRUE)
crossval_8pc <- table(cvagroups, cvaresults_8pc_CV$class) # confusion matrix of specimen assignments
sum(diag(prop.table(crossval_8pc))) # total percent correct

# Supplementary Table 4
altcrossval_8pc <- rbind(crossval_8pc[1,]/sum(crossval_8pc[1,]), crossval_8pc[2,]/sum(crossval_8pc[2,]), crossval_8pc[3,]/sum(crossval_8pc[3,]), crossval_8pc[4,]/sum(crossval_8pc[4,]))
dimnames(altcrossval_8pc) <- list("Actual"=c("F", "HC", "LC", "O"), "Predicted (CV)"=c("F", "HC", "LC", "O"))
print(round(altcrossval_8pc, 3))
```

#### Supplementary Figure 3 (comparing 30PC vs 8PC CVAs):

```{r, echo = F, fig.width=9.5, fig.height=7}
# CV1 vs. CV2 - 30PC model
hull_cva_12_30pc <- C_30pc_info %>% group_by(Habitat) %>%
  slice(chull(CV1, CV2))
cva_1v2_30pc <- ggplot()+
  geom_polygon(data = hull_cva_12_30pc, aes(x = CV1, y = CV2, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ # plot convex hulls
  scale_fill_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_point(data = C_30pc_info, mapping = aes(x = CV1, y = CV2, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  tribe_shapes() +
  theme_classic(base_size = 22) +
  xlab("CV1")+
  ylab("CV2")

# CV1 vs. CV3 - 30 PC model
hull_cva_13_30pc <- C_30pc_info %>% group_by(Habitat) %>%
  slice(chull(CV1, CV3))
cva_1v3_30pc <- ggplot()+
  geom_polygon(data = hull_cva_13_30pc, aes(x = CV1, y = CV3, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ # plot convex hulls
  scale_fill_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_point(data = C_30pc_info, mapping = aes(x = CV1, y = CV3, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  tribe_shapes() +
  theme_classic(base_size = 22) +
  xlab("CV1")+
  ylab("CV3")

# show plots
cva_1v2_30pc
cva_1v3_30pc
```

```{r, echo = F, fig.width=9.5, fig.height=7}
# CV1 vs. CV2 - 8 PC model
hull_cva_12_8pc <- C_8pc_info %>% group_by(Habitat) %>%
  slice(chull(CV1, CV2))
cva_1v2_8pc <- ggplot()+
  geom_polygon(data = hull_cva_12_8pc, aes(x = CV1, y = CV2, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ # plot convex hulls
  scale_fill_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_point(data = C_8pc_info, mapping = aes(x = CV1, y = CV2, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  tribe_shapes() +
  theme_classic(base_size = 22) +
  xlab("CV1")+
  ylab("CV2")

# CV1 vs. CV3 - 8 PC model
hull_cva_13_8pc <- C_8pc_info %>% group_by(Habitat) %>%
  slice(chull(CV1, CV3))
cva_1v3_8pc <- ggplot()+
  geom_polygon(data = hull_cva_13_8pc, aes(x = CV1, y = CV3, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ # plot convex hulls
  scale_fill_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  geom_point(data = C_8pc_info, mapping = aes(x = CV1, y = CV3, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1, labels = c("Forest", "Heavy Cover", "Light Cover", "Open"))+
  tribe_shapes() +
  theme_classic(base_size = 22) +
  xlab("CV1")+
  ylab("CV3")

# show plots
cva_1v2_8pc
cva_1v3_8pc
```

Combine plots:
```{r, fig.width=9.5, fig.height=15}
( (cva_1v2_30pc / cva_1v3_30pc ) | (cva_1v2_8pc / cva_1v3_8pc) ) + plot_layout(guides = "collect")
```

### Between-group PCA as an alternative to CVA

Mitteroecker & Bookstein (2011) state that in the case where the number of variables exceeds the number of cases, CVA should be avoided in favor of methods that do not require inversion of a covariance matrix, such as between-group PCA.
```{r}
# morpho package function
bg_pca <- groupPCA(dataarray = GPA_all$coords[,,1:136],
                   groups = cvagroups,
                   cv = TRUE)
bg_pca
```

Are those differences between groups significant? Procrustes ANOVA (shape ~ habitat):
```{r}
# regress species mean Procrustes shapes against habitat to identify if group means are significantly different (recommendation of Cardini & Polly 2020)
summary(procD.lm(mean_shapes ~ species_info$Habitat))
```

Yes, habitat is significantly associated with species mean shape according to a Procrustes ANOVA. The R2 is 0.13, with a p-value of 0.014. 

#### Figure 5: bgPC1 vs. bgPC2 & bgPC1 vs. bgPC3
```{r}
# generate data frame for plotting
bg_pca_scores <- as.data.frame(bg_pca[["Scores"]])
bg_pca_scores <- cbind(bg_pca_scores, cvagroups, modern_astragali_info$Tribe)
colnames(bg_pca_scores) <- c("Dim1", "Dim2", "Dim3", "Habitat", "Tribe")

# plot the scores
hull_bgpca_12 <- bg_pca_scores %>% group_by(Habitat) %>%
  slice(chull(Dim1, Dim2))
bgpca_1v2 <- ggplot() +
  geom_polygon(data = hull_bgpca_12, aes(x = Dim1, y = Dim2, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ 
  geom_point(data = bg_pca_scores, aes(x = Dim1, y = Dim2, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1)+
  scale_fill_brewer(palette = "BrBG", direction = -1)+
  tribe_shapes()+
  xlab("between-group PC1 (74.9%)")+
  ylab("between-group PC2 (14.2%)")+
  theme_classic(base_size = 22)

hull_bgpca_13 <- bg_pca_scores %>% group_by(Habitat) %>%
  slice(chull(Dim1, Dim3))
bgpca_1v3 <- ggplot() +
  geom_polygon(data = hull_bgpca_13, aes(x = Dim1, y = Dim3, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ 
  geom_point(data = bg_pca_scores, aes(x = Dim1, y = Dim3, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1)+
  scale_fill_brewer(palette = "BrBG", direction = -1)+
  tribe_shapes()+
  xlab("between-group PC1 (74.9%)")+
  ylab("between-group PC3 (10.9%)")+
  theme_classic(base_size = 22)

hull_bgpca_23 <- bg_pca_scores %>% group_by(Habitat) %>%
  slice(chull(Dim2, Dim3))
bgpca_2v3 <- ggplot() +
  geom_polygon(data = hull_bgpca_23, aes(x = Dim2, y = Dim3, color = Habitat, fill = Habitat), linewidth = 1, alpha = 0.2)+ 
  geom_point(data = bg_pca_scores, aes(x = Dim2, y = Dim3, color = Habitat, shape = Tribe), size = 3)+
  scale_color_brewer(palette = "BrBG", direction = -1)+
  scale_fill_brewer(palette = "BrBG", direction = -1)+
  tribe_shapes()+
  xlab("between-group PC2 (14.2%)")+
  ylab("between-group PC3 (10.9%)")+
  theme_classic(base_size = 22)
```

```{r, fig.height=6, fig.width=8}
(bgpca_1v2 / bgpca_1v3) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

#### Figure 6: warps of max and min values along bgPC axes

```{r}
# generate mesh of mean shape by warping specimen closest to the mean to the mean shape from the GPA
ref_spec <- mshape(GPA_all$coords[,,1:136]) # find mean shape of modern astragali
mean_spec <- findMeanSpec(GPA_all$coords[,,1:136]) # actual specimen closest to mean
mean_filename <- modern_astragali_info$filename[mean_spec] # filename of mean specimen
mean_filename <- gsub("\\..*",'', mean_filename) # removes the ".csv" from the end of the filename (removes everything after the period)

# read in .ply file of mean specimen
mean_mesh <- read.ply(file = paste("data/",mean_filename,".ply", sep = ""), ShowSpecimen = TRUE)

# non-Procrustes landmarks for actual specimen closest to the mean
mean_landmarks <- all_landmarks[,,mean_spec]

# warp real specimen closest to the mean to the mean shape
mean_warp <- warpRefMesh(mesh = mean_mesh, mesh.coord = mean_landmarks, ref = ref_spec, centered = F)
plot3d(mean_warp, aspect = F) 

# get landmarks from max and min along each bg-PC
bgPC1_max <- restoreShapes(scores = max(bg_pca$Scores[,1]), PC = bg_pca$groupPCs[,1], mshape = ref_spec)
bgPC1_min <- restoreShapes(scores = min(bg_pca$Scores[,1]), PC = bg_pca$groupPCs[,1], mshape = ref_spec)

bgPC2_max <- restoreShapes(scores = max(bg_pca$Scores[,2]), PC = bg_pca$groupPCs[,2], mshape = ref_spec)
bgPC2_min <- restoreShapes(scores = min(bg_pca$Scores[,2]), PC = bg_pca$groupPCs[,2], mshape = ref_spec)

bgPC3_max <- restoreShapes(scores = max(bg_pca$Scores[,3]), PC = bg_pca$groupPCs[,3], mshape = ref_spec)
bgPC3_min <- restoreShapes(scores = min(bg_pca$Scores[,3]), PC = bg_pca$groupPCs[,3], mshape = ref_spec)

# bgPC1 max warp 
warp_bgpc1_max <- plotRefToTarget(ref_spec, bgPC1_max, mesh = mean_warp, method = "surface")
meshDist(x = warp_bgpc1_max, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)
# bgPC1 min warp 
warp_bgpc1_min <- plotRefToTarget(ref_spec, bgPC1_min, mesh = mean_warp, method = "surface")
meshDist(x = warp_bgpc1_min, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)

# bgPC2 max warp 
warp_bgpc2_max <- plotRefToTarget(ref_spec, bgPC2_max, mesh = mean_warp, method = "surface")
meshDist(x = warp_bgpc2_max, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)
# bgPC2 min warp 
warp_bgpc2_min <- plotRefToTarget(ref_spec, bgPC2_min, mesh = mean_warp, method = "surface")
meshDist(x = warp_bgpc2_min, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)

# bgPC3 max warp 
warp_bgpc3_max <- plotRefToTarget(ref_spec, bgPC3_max, mesh = mean_warp, method = "surface")
meshDist(x = warp_bgpc3_max, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)
# bgPC3 min warp 
warp_bgpc3_min <- plotRefToTarget(ref_spec, bgPC3_min, mesh = mean_warp, method = "surface")
meshDist(x = warp_bgpc3_min, mesh2 = mean_warp, rampcolors = c("maroon", "white", "navy"), save = F)
```

#### Misclassifications within the modern sample:
```{r}
bgpca_predictions <- classify(bg_pca, cv=T)
bgpca_predictions <- bgpca_predictions$class
bgpca_predictions <- bind_cols(modern_astragali_info[,c("Tribe", "Genus_species", "Habitat")], bgpca_predictions)
colnames(bgpca_predictions)[4] <- "pred_habitat"

# which predictions are correct (TRUE) versus incorrect (FALSE)
bgpca_predictions$match <- (bgpca_predictions$Habitat == bgpca_predictions$pred_habitat)

# proportions of correct (TRUE) and incorrect (FALSE) classifications by tribe
tribe_prediction_performance_bgpca <- table(bgpca_predictions$Tribe, bgpca_predictions$match)
tribe_prediction_performance_prop_bgpca <- prop.table(tribe_prediction_performance_bgpca, 1)

# proportions of correct (TRUE) and incorrect (FALSE) classifications by species
species_prediction_performance_bgpca <- table(bgpca_predictions$Genus_species, bgpca_predictions$match)
species_prediction_performance_prop_bgpca <- prop.table(species_prediction_performance_bgpca, 1)
```

### Predict fossils using CVA

First, project fossil Procrustes coordinates into PC space generated from modern data:
```{r}
# subset matrix of fossil Procrustes coordinates
fossil_procrustes <- GPA_all$coords[,,137:dim(GPA_all$coords)[3]]

# convert Procrustes coordinates to wide format for use outside geomorph
all_coords2d <- two.d.array(GPA_all$coords)
consensusvec <- apply(all_coords2d, 2, mean) # GPA consensus in vector format 

# calculate residuals of fossil specimens
consensus <- GPA_all$consensus
fossil_resids <- sweep(fossil_procrustes, c(1,2), consensus) # subtract mean shape from all fossils

# extract rotation matrix from modern PCA
rotation <- PCA_modern$rotation

# calculate PC scores for fossils by multiplying matrices
fossil_pc_scores <- two.d.array(fossil_resids)%*%rotation

# turn it into a data frame
fossil_pc_scores <- as.data.frame(fossil_pc_scores)
fossil_pc_scores$Tribe <- "Fossil"
```

Next, conduct CVA on modern data without scaling of PC scores--scaling PCs including fossils will make CV axis estimation dependent on fossil data, which we want to avoid. Using the un-scaled PCs rather than scaled provides the same results for modern data, and demonstrates that relationships/relative positions among data points don't change, only the specific values along each axis.
```{r}
# conduct the CVA (without cross-validation) on all un-scaled PC scores:
cvaresults <- lda(mod_PC_scores, cvagroups, CV=FALSE)
A <- cvaresults$scaling
grpmeans <- cvaresults$means

# project modern specimens onto CVA
C <- mod_PC_scores%*%A 

# add specimen info to CV scores for plotting
C_modern <- data.frame(CV1 = C[,1],
                     CV2 = C[,2],
                     CV3 = C[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe)

# predict fossil scores
C_fossil <- predict(cvaresults, newdata = fossil_pc_scores[,1:53])
fossil_scores <- C_fossil$x # CV scores for all fossils
post_prob_fossils<- C_fossil$posterior # probabilities of group membership

# make this easier to visualize
C_modern_info <- data.frame(CV1 = C_modern[,1],
                     CV2 = C_modern[,2],
                     CV3 = C_modern[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe,
                     Specimen = modern_astragali_info$Spec.number)
fossil_scores_info <- data.frame(CV1 = fossil_scores[,1],
                            CV2 = fossil_scores[,2],
                            CV3 = fossil_scores[,3],
                            Locality = fossil_astragali_info$Community, 
                            Tribe = "Fossil",
                            Specimen = fossil_astragali_info$Spec.number,
                            Submember = fossil_astragali_info$submember,
                            Submember_2 = fossil_astragali_info$submember_2)

fossil_scores_info <- bind_cols(fossil_scores_info, post_prob_fossils)

# record the category that the DFA predicted as most likely for each individual
fossil_scores_info$likely_category <- NA
fossil_scores_info$likely_category <- colnames(fossil_scores_info[,9:12])[apply(fossil_scores_info[,9:12], 1, which.max)]

fossil_scores_info %>% group_by(Submember_2, likely_category) %>%
  tally()
```

Repeat this with the CVA using the first 30PCs:
```{r}
# conduct the CVA (without cross-validation) on the first 30 un-scaled PC scores:
cvaresults_30pc <- lda(mod_PC_scores[,1:30], cvagroups, CV=FALSE)
A_30pc <- cvaresults_30pc$scaling
grpmeans_30pc <- cvaresults_30pc$means

# project modern specimens onto CVA
C_30pc <- mod_PC_scores[,1:30]%*%A_30pc

# add specimen info to CV scores for plotting
C_modern_30pc <- data.frame(CV1 = C_30pc[,1],
                     CV2 = C_30pc[,2],
                     CV3 = C_30pc[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe)

# predict fossil scores
C_fossil_30pc <- predict(cvaresults_30pc, newdata = fossil_pc_scores[,1:30])
fossil_scores_30pc <- C_fossil_30pc$x # CV scores for all fossils
post_prob_fossils_30pc<- C_fossil_30pc$posterior # probabilities of group membership

# make this easier to visualize
C_modern_info_30pc <- data.frame(CV1 = C_modern_30pc[,1],
                     CV2 = C_modern_30pc[,2],
                     CV3 = C_modern_30pc[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe,
                     Specimen = modern_astragali_info$Spec.number)
fossil_scores_info_30pc <- data.frame(CV1 = fossil_scores_30pc[,1],
                            CV2 = fossil_scores_30pc[,2],
                            CV3 = fossil_scores_30pc[,3],
                            Locality = fossil_astragali_info$Community, 
                            Tribe = "Fossil",
                            Specimen = fossil_astragali_info$Spec.number,
                            Submember = fossil_astragali_info$submember,
                            Submember_2 = fossil_astragali_info$submember_2)

fossil_scores_info_30pc <- bind_cols(fossil_scores_info_30pc, post_prob_fossils_30pc)

# record the category that the CVA predicted as most likely for each individual
fossil_scores_info_30pc$likely_category <- NA
fossil_scores_info_30pc$likely_category <- colnames(fossil_scores_info_30pc[,9:12])[apply(fossil_scores_info_30pc[,9:12], 1, which.max)]

fossil_scores_info_30pc %>% group_by(Submember_2, likely_category) %>%
  tally()
```

Repeat this with the CVA using PCs 1-8 only:
```{r}
# conduct the CVA (without cross-validation) on the first 8 un-scaled PC scores:
cvaresults_8pc <- lda(mod_PC_scores[,1:8], cvagroups, CV=FALSE)
A_8pc <- cvaresults_8pc$scaling
grpmeans_8pc <- cvaresults_8pc$means

# project modern specimens onto CVA
C_8pc <- mod_PC_scores[,1:8]%*%A_8pc

# add specimen info to CV scores for plotting
C_modern_8pc <- data.frame(CV1 = C_8pc[,1],
                     CV2 = C_8pc[,2],
                     CV3 = C_8pc[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe)

# predict fossil scores
C_fossil_8pc <- predict(cvaresults_8pc, newdata = fossil_pc_scores[,1:8])
fossil_scores_8pc <- C_fossil_8pc$x # CV scores for all fossils
post_prob_fossils_8pc<- C_fossil_8pc$posterior # probabilities of group membership

# make this easier to visualize
C_modern_info_8pc <- data.frame(CV1 = C_modern_8pc[,1],
                     CV2 = C_modern_8pc[,2],
                     CV3 = C_modern_8pc[,3],
                     Habitat = modern_astragali_info$Habitat,
                     Tribe = modern_astragali_info$Tribe,
                     Specimen = modern_astragali_info$Spec.number)
fossil_scores_info_8pc <- data.frame(CV1 = fossil_scores_8pc[,1],
                            CV2 = fossil_scores_8pc[,2],
                            CV3 = fossil_scores_8pc[,3],
                            Locality = fossil_astragali_info$Community, 
                            Tribe = "Fossil",
                            Specimen = fossil_astragali_info$Spec.number,
                            Submember = fossil_astragali_info$submember,
                            Submember_2 = fossil_astragali_info$submember_2)

fossil_scores_info_8pc <- bind_cols(fossil_scores_info_8pc, post_prob_fossils_8pc)

# record the category that the DFA predicted as most likely for each individual
fossil_scores_info_8pc$likely_category <- NA
fossil_scores_info_8pc$likely_category <- colnames(fossil_scores_info_8pc[,9:12])[apply(fossil_scores_info_8pc[,9:12], 1, which.max)]

fossil_scores_info_8pc %>% group_by(Submember_2, likely_category) %>%
  tally()
```

### Predict fossils using the bgPCA
```{r}
# project fossils onto the bgPCA space
bg_pca_pred <- as.data.frame(predict(bg_pca, newdata = GPA_all$coords[,,137:dim(GPA_all$coords)[3]])) 
colnames(bg_pca_pred) <- c("Dim1", "Dim2", "Dim3")

# classify fossils using cross-validated bgPCA
bg_pca_classif <- classify(bg_pca, newdata = GPA_all$coords[,,137:dim(GPA_all$coords)[3]], cv = TRUE)

# generate dataframe for plotting
fossil_scores_bgpca <- bind_cols(bg_pca_pred, fossil_astragali_info[c("Spec.number", "Community", "submember", "submember_2")])
fossil_scores_bgpca$pred_habitat <- bg_pca_classif$class
```

### Figure 7: Lower vs. Upper G

CVA using all 53 PCs:
```{r, fig.width = 10.5, fig.height = 7}
# delineate upper versus lower G in dataset
fossil_scores_info$lower <- NA
fossil_scores_info[fossil_scores_info$Submember %in% c("G-15","G-19","G-24","G-27","G-28"),]$lower <- F # upper G
fossil_scores_info[!fossil_scores_info$Submember %in% c("G-15","G-19","G-24","G-27","G-28"),]$lower <- T # lower G

# chi-square to test for significant differences in habitat categories between upper and lower G
upper_v_lower <- table(fossil_scores_info$lower, fossil_scores_info$likely_category)
dimnames(upper_v_lower)[[1]] <- c("upper", "lower")
set.seed(1212)
chisq.test(upper_v_lower, simulate.p.value = T)

# upper versus lower G (G1-13 versus G15-29) -- this plot includes everything
upper_v_lower_plot <- fossil_scores_info %>% group_by(lower, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = lower, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  scale_y_discrete(limits = rev, labels = c("G-1-13\n2.26 -\n2.08 Ma", "G-15-29\n2.07 -\n1.92 Ma"))+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab(NULL)+
  annotate(geom = "text", x = c(1.03, 1.03), y = c(1,2), label = c("n =\n164", "n =\n31"), hjust = 0.2, size = 5)+
  theme_classic(base_size = 22)+
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 0.5))
upper_v_lower_plot
```

bgPCA predictions:
```{r,fig.width = 10.5, fig.height = 7}
# delineate upper versus lower G in dataset
fossil_scores_bgpca$lower <- NA
fossil_scores_bgpca[fossil_scores_bgpca$submember %in% c("G-15","G-19","G-24","G-27","G-28"),]$lower <- F # upper G
fossil_scores_bgpca[!fossil_scores_bgpca$submember %in% c("G-15","G-19","G-24","G-27","G-28"),]$lower <- T # lower G

# chi-square to test for significant differences in habitat categories between upper and lower G
upper_v_lower_bgpca <- table(fossil_scores_bgpca$lower, fossil_scores_bgpca$pred_habitat)
dimnames(upper_v_lower_bgpca)[[1]] <- c("upper", "lower")
set.seed(1212)
chisq.test(upper_v_lower_bgpca, simulate.p.value = T)

# upper versus lower G (G1-13 versus G15-29) -- this plot includes everything
upper_v_lower_plot_bgpca <- fossil_scores_bgpca %>% group_by(lower, pred_habitat) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = lower, fill = pred_habitat), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  scale_y_discrete(limits = rev, labels = c("G-1-13\n2.26 -\n2.08 Ma", "G-15-29\n2.07 -\n1.92 Ma"))+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab(NULL)+
  annotate(geom = "text", x = c(1.03, 1.03), y = c(1,2), label = c("n =\n164", "n =\n31"), hjust = 0.2, size = 5)+
  theme_classic(base_size = 22)+
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 0.5))

upper_v_lower_plot_bgpca
```

Combine the two plots:
```{r, fig.height=12, fig.width=10.5}
upper_v_lower_plot / upper_v_lower_plot_bgpca + plot_layout(guides = "collect", axis_titles = "collect_x") & theme(legend.position = "bottom")
```

### Supplementary Figure 4: CVAs using 30 PCs and 8PCs, Lower vs. Upper G

CVA using 30 PCs:
```{r, fig.width = 10.5, fig.height = 7}
# delineate upper versus lower G in dataset
fossil_scores_info_30pc$lower <- NA
fossil_scores_info_30pc[fossil_scores_info_30pc$Submember %in% c("G-15","G-19","G-24","G-27","G-28"),]$lower <- F # upper G
fossil_scores_info_30pc[!fossil_scores_info_30pc$Submember %in% c("G-15","G-19","G-24","G-27","G-28"),]$lower <- T # lower G

# chi-square to test for significant differences in habitat categories between upper and lower G
upper_v_lower_30pc <- table(fossil_scores_info_30pc$lower, fossil_scores_info_30pc$likely_category)
dimnames(upper_v_lower_30pc)[[1]] <- c("upper", "lower")
set.seed(1212)
chisq.test(upper_v_lower_30pc, simulate.p.value = T)

# upper versus lower G (G1-13 versus G15-29) -- this plot includes everything
upper_v_lower_plot_30pc <- fossil_scores_info_30pc %>% group_by(lower, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = lower, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  scale_y_discrete(limits = rev, labels = c("G-1-13\n2.26 -\n2.08 Ma", "G-15-29\n2.07 -\n1.92 Ma"))+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab(NULL)+
  annotate(geom = "text", x = c(1.03, 1.03), y = c(1,2), label = c("n =\n164", "n =\n31"), hjust = 0.2, size = 5)+
  theme_classic(base_size = 22)+
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 0.5))
upper_v_lower_plot_30pc
```

CVA using 8 PCs:
```{r, fig.width = 10.5, fig.height = 7}
# delineate upper versus lower G in dataset
fossil_scores_info_8pc$lower <- NA
fossil_scores_info_8pc[fossil_scores_info_8pc$Submember %in% c("G-15","G-19","G-24","G-27","G-28"),]$lower <- F # upper G
fossil_scores_info_8pc[!fossil_scores_info_8pc$Submember %in% c("G-15","G-19","G-24","G-27","G-28"),]$lower <- T # lower G

# chi-square to test for significant differences in habitat categories between upper and lower G
upper_v_lower_8pc <- table(fossil_scores_info_8pc$lower, fossil_scores_info_8pc$likely_category)
dimnames(upper_v_lower_8pc)[[1]] <- c("upper", "lower")
set.seed(1212)
chisq.test(upper_v_lower_8pc, simulate.p.value = T)

# upper versus lower G (G1-13 versus G15-29) -- this plot includes everything
upper_v_lower_plot_8pc <- fossil_scores_info_8pc %>% group_by(lower, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = lower, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  scale_y_discrete(limits = rev, labels = c("G-1-13\n2.26 -\n2.08 Ma", "G-15-29\n2.07 -\n1.92 Ma"))+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab(NULL)+
  annotate(geom = "text", x = c(1.03, 1.03), y = c(1,2), label = c("n =\n164", "n =\n31"), hjust = 0.2, size = 5)+
  theme_classic(base_size = 22)+
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 0.5))
upper_v_lower_plot_8pc
```

Combine plots:
```{r, fig.height=12, fig.width=8}
upper_v_lower_plot_30pc / upper_v_lower_plot_8pc +
  plot_layout(guides = "collect", axis_titles = "collect_x" ) & theme(legend.position = "bottom")
```

### Supplementary Figure 5: All submembers, CVA vs bgPCA
Showing all of the sub-members, even ones with wide temporal ranges (e.g., G-1-9) and ones represented by 2 or fewer fossils. Supplementary Figure 5 is the CVA with all PCs versus the bgPCA predictions. 

CVA with all PCs:
```{r, fig.width = 9, fig.height = 11}
# plot everything with the most detailed sub-member info
sample_sizes <- as.vector(fossil_scores_info %>% group_by(Submember_2) %>% tally())
sample_sizes <- sample_sizes[["n"]]
sample_sizes <- paste("n =", sample_sizes)

submember_2_plot <- fossil_scores_info %>% group_by(Submember_2, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = Submember_2, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab("Submember")+
  annotate(geom = "text", x = 1.03, y = as.vector(1:length(sample_sizes)), label = sample_sizes, hjust = 0.2, size = 4)+
  theme(legend.position = "bottom")

submember_2_plot
```

bgPCA predictions:
```{r, fig.width = 9, fig.height = 11}
# plot everything with the most detailed sub-member info
submember_2_plot_bgpca <- fossil_scores_bgpca %>% group_by(submember_2, pred_habitat) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = submember_2, fill = pred_habitat), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab("Submember")+
  annotate(geom = "text", x = 1.03, y = as.vector(1:length(sample_sizes)), label = sample_sizes, hjust = 0.2, size = 4)+
  theme(legend.position = "bottom")

submember_2_plot_bgpca
```

Combine plots:
```{r, fig.height=10, fig.width=18}
(submember_2_plot | submember_2_plot_bgpca) + plot_layout(axis_titles = "collect", axes = "collect_y", guides = "collect") & theme(legend.position = "bottom") 
```

### Supplementary Figure 6: All submembers, 30 PC CVA vs 8 PC CVA
Same as Supplementary Figure 5, but for the CVAs using 30 PCs and 8 PCs.

30 PC CVA predictions:
```{r, fig.width = 9, fig.height = 11}
submember_2_plot_30pc <- fossil_scores_info_30pc %>% group_by(Submember_2, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = Submember_2, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab("Submember")+
  annotate(geom = "text", x = 1.03, y = as.vector(1:length(sample_sizes)), label = sample_sizes, hjust = 0.2, size = 4)+
  theme(legend.position = "bottom")

submember_2_plot_30pc
```

8 PC CVA predictions:
```{r, fig.width = 9, fig.height = 11}
submember_2_plot_8pc <- fossil_scores_info_8pc %>% group_by(Submember_2, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = Submember_2, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab("Submember")+
  annotate(geom = "text", x = 1.03, y = as.vector(1:length(sample_sizes)), label = sample_sizes, hjust = 0.2, size = 4)+
  theme(legend.position = "bottom")

submember_2_plot_8pc
```

Combine plots:
```{r fig.height=12, fig.width=18}
(submember_2_plot_30pc | submember_2_plot_8pc) + plot_layout(axis_titles = "collect", axes = "collect_y", guides = "collect") & theme(legend.position = "bottom") 
``` 

### Figure 8: Well-sampled and narrowly temporally constrained intervals

Main text figure 8 includes only the CVA using all PCs and the bgPCA.

Predictions from CVA using all PCs:
```{r, fig.width = 9, fig.height = 9}
# reorder factor levels so that the submembers are in chronological order from oldest to youngest
fossil_scores_info$Submember <- as.factor(fossil_scores_info$Submember)
levels(fossil_scores_info$Submember)
fossil_scores_info$Submember <- factor(fossil_scores_info$Submember, levels = c("G-1","G-2","G-3","G-4","G-5","G-6","G-7","G-8","G-9","G-10","G-11","G-12","G-13","G-15","G-19","G-24","G-27","G-28"))

fossil_scores_info$Submember_2 <- as.factor(fossil_scores_info$Submember_2)
levels(fossil_scores_info$Submember_2)
fossil_scores_info$Submember_2 <- factor(fossil_scores_info$Submember_2, levels = c("G-1", "G-1-5", "G-1-9", "G-1-13", "G-3", "G-3-8", "G-3-13", "G-4", "G-4-13", "G-5", "G-6", "G-6-8", "G-7", "G-7-8", "G-8", "G-8-9", "G-9", "G-10-11", "G-11", "G-12", "G-12-13", "G-13", "G-15", "G-15-29", "G-19", "G-19-20", "G-24", "G-27", "G-27-28", "G-28-29"))

# only n > 2 and combining some groups (adjacent sub-members) 
fossil_scores_info$plotting <- NA
fossil_scores_info$plotting <- fossil_scores_info$Submember_2
fossil_scores_info[fossil_scores_info$plotting %in% c("G-27", "G-27-28"),]$plotting <- "G-27-28"
fossil_scores_info[fossil_scores_info$plotting %in% c("G-19", "G-19-20"),]$plotting <- "G-19-20"
fossil_scores_info[fossil_scores_info$plotting %in% c("G-10-11", "G-11"),]$plotting <- "G-10-11"
fossil_scores_info[fossil_scores_info$plotting %in% c("G-8-9", "G-9"),]$plotting <- "G-8-9"

# sample sizes for plot
subset_samples <- as.vector(fossil_scores_info %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) %>% group_by(plotting) %>% tally())
subset_samples <- subset_samples[["n"]]
subset_samples <- paste("n =", subset_samples)

# chi-squared
subset <- fossil_scores_info %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) 
subset$plotting <- factor(subset$plotting, levels = c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29"))
subset <- table(subset$plotting, subset$likely_category)
set.seed(1212)
chisq.test(subset, simulate.p.value = T)

# plot
submember_plot_clean <- fossil_scores_info %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) %>% group_by(plotting, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = plotting, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab("Submember")+
  annotate(geom = "text", x = 1.03, y = as.vector(1:(length(subset_samples))), label = subset_samples, hjust = 0.2, size = 4)+
  theme(legend.position = "bottom")

submember_plot_clean
```

Predictions from bgPCA:
```{r, fig.width = 9, fig.height = 9}
# reorder factor levels so that the submembers are in chronological order from oldest to youngest
fossil_scores_bgpca$submember <- as.factor(fossil_scores_bgpca$submember)
levels(fossil_scores_bgpca$submember)
fossil_scores_bgpca$submember <- factor(fossil_scores_bgpca$submember, levels = c("G-1","G-2","G-3","G-4","G-5","G-6","G-7","G-8","G-9","G-10","G-11","G-12","G-13","G-15","G-19","G-24","G-27","G-28"))

fossil_scores_bgpca$submember_2 <- as.factor(fossil_scores_bgpca$submember_2)
levels(fossil_scores_bgpca$submember_2)
fossil_scores_bgpca$submember_2 <- factor(fossil_scores_bgpca$submember_2, levels = c("G-1", "G-1-5", "G-1-9", "G-1-13", "G-3", "G-3-8", "G-3-13", "G-4", "G-4-13", "G-5", "G-6", "G-6-8", "G-7", "G-7-8", "G-8", "G-8-9", "G-9", "G-10-11", "G-11", "G-12", "G-12-13", "G-13", "G-15", "G-15-29", "G-19", "G-19-20", "G-24", "G-27", "G-27-28", "G-28-29"))

# only n > 2 and combining some groups (adjacent sub-members) 
fossil_scores_bgpca$plotting <- NA
fossil_scores_bgpca$plotting <- fossil_scores_bgpca$submember_2
fossil_scores_bgpca[fossil_scores_bgpca$plotting %in% c("G-27", "G-27-28"),]$plotting <- "G-27-28"
fossil_scores_bgpca[fossil_scores_bgpca$plotting %in% c("G-19", "G-19-20"),]$plotting <- "G-19-20"
fossil_scores_bgpca[fossil_scores_bgpca$plotting %in% c("G-10-11", "G-11"),]$plotting <- "G-10-11"
fossil_scores_bgpca[fossil_scores_bgpca$plotting %in% c("G-8-9", "G-9"),]$plotting <- "G-8-9"

# chi-squared
subset_bgpca <- fossil_scores_bgpca %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) 
subset_bgpca$plotting <- factor(subset_bgpca$plotting, levels = c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29"))
subset_bgpca <- table(subset_bgpca$plotting, subset_bgpca$pred_habitat)
set.seed(1212)
chisq.test(subset_bgpca, simulate.p.value = T)

# plot
submember_plot_clean_bgpca <- fossil_scores_bgpca %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) %>% group_by(plotting, pred_habitat) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = plotting, fill = pred_habitat), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab("Submember")+
  annotate(geom = "text", x = 1.03, y = as.vector(1:(length(subset_samples))), label = subset_samples, hjust = 0.2, size = 4)+
  theme(legend.position = "bottom")

submember_plot_clean_bgpca
```

Compare to CVA with full shape:
```{r, fig.height=9, fig.width=18}
(submember_plot_clean | submember_plot_clean_bgpca) + plot_layout(axis_titles = "collect", axes = "collect_y", guides = "collect") & theme(legend.position = "bottom") 
```


### Supplementary Figure 7: well-sampled and narrow temporal intervals, 30 PC vs 8 PC CVAs
Well-sampled and better temporally constrained intervals for the 30 PC and 8 PC CVAs.

Predictions from 30 PC CVA:
```{r, fig.width = 9, fig.height = 9}
# reorder factor levels so that the submembers are in chronological order from oldest to youngest
fossil_scores_info_30pc$Submember <- as.factor(fossil_scores_info_30pc$Submember)
levels(fossil_scores_info_30pc$Submember)
fossil_scores_info_30pc$Submember <- factor(fossil_scores_info_30pc$Submember, levels = c("G-1","G-2","G-3","G-4","G-5","G-6","G-7","G-8","G-9","G-10","G-11","G-12","G-13","G-15","G-19","G-24","G-27","G-28"))

fossil_scores_info_30pc$Submember_2 <- as.factor(fossil_scores_info_30pc$Submember_2)
levels(fossil_scores_info_30pc$Submember_2)
fossil_scores_info_30pc$Submember_2 <- factor(fossil_scores_info_30pc$Submember_2, levels = c("G-1", "G-1-5", "G-1-9", "G-1-13", "G-3", "G-3-8", "G-3-13", "G-4", "G-4-13", "G-5", "G-6", "G-6-8", "G-7", "G-7-8", "G-8", "G-8-9", "G-9", "G-10-11", "G-11", "G-12", "G-12-13", "G-13", "G-15", "G-15-29", "G-19", "G-19-20", "G-24", "G-27", "G-27-28", "G-28-29"))

# only n > 2 and combining some groups (adjacent sub-members) 
fossil_scores_info_30pc$plotting <- NA
fossil_scores_info_30pc$plotting <- fossil_scores_info_30pc$Submember_2
fossil_scores_info_30pc[fossil_scores_info_30pc$plotting %in% c("G-27", "G-27-28"),]$plotting <- "G-27-28"
fossil_scores_info_30pc[fossil_scores_info_30pc$plotting %in% c("G-19", "G-19-20"),]$plotting <- "G-19-20"
fossil_scores_info_30pc[fossil_scores_info_30pc$plotting %in% c("G-10-11", "G-11"),]$plotting <- "G-10-11"
fossil_scores_info_30pc[fossil_scores_info_30pc$plotting %in% c("G-8-9", "G-9"),]$plotting <- "G-8-9"

# chi-squared
subset_30pc <- fossil_scores_info_30pc %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) 
subset_30pc$plotting <- factor(subset_30pc$plotting, levels = c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29"))
subset_30pc <- table(subset_30pc$plotting, subset_30pc$likely_category)
set.seed(1212)
chisq.test(subset_30pc, simulate.p.value = T)

# plot
submember_plot_clean_30pc <- fossil_scores_info_30pc %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) %>% group_by(plotting, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = plotting, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab("Submember")+
  annotate(geom = "text", x = 1.03, y = as.vector(1:(length(subset_samples))), label = subset_samples, hjust = 0.2, size = 4)+
  theme(legend.position = "bottom")

submember_plot_clean_30pc
```

Predictions from 8 PC CVA:
```{r, fig.width = 9, fig.height = 9}
# reorder factor levels so that the submembers are in chronological order from oldest to youngest
fossil_scores_info_8pc$Submember <- as.factor(fossil_scores_info_8pc$Submember)
levels(fossil_scores_info_8pc$Submember)
fossil_scores_info_8pc$Submember <- factor(fossil_scores_info_8pc$Submember, levels = c("G-1","G-2","G-3","G-4","G-5","G-6","G-7","G-8","G-9","G-10","G-11","G-12","G-13","G-15","G-19","G-24","G-27","G-28"))

fossil_scores_info_8pc$Submember_2 <- as.factor(fossil_scores_info_8pc$Submember_2)
levels(fossil_scores_info_8pc$Submember_2)
fossil_scores_info_8pc$Submember_2 <- factor(fossil_scores_info_8pc$Submember_2, levels = c("G-1", "G-1-5", "G-1-9", "G-1-13", "G-3", "G-3-8", "G-3-13", "G-4", "G-4-13", "G-5", "G-6", "G-6-8", "G-7", "G-7-8", "G-8", "G-8-9", "G-9", "G-10-11", "G-11", "G-12", "G-12-13", "G-13", "G-15", "G-15-29", "G-19", "G-19-20", "G-24", "G-27", "G-27-28", "G-28-29"))

# only n > 2 and combining some groups (adjacent sub-members) 
fossil_scores_info_8pc$plotting <- NA
fossil_scores_info_8pc$plotting <- fossil_scores_info_8pc$Submember_2
fossil_scores_info_8pc[fossil_scores_info_8pc$plotting %in% c("G-27", "G-27-28"),]$plotting <- "G-27-28"
fossil_scores_info_8pc[fossil_scores_info_8pc$plotting %in% c("G-19", "G-19-20"),]$plotting <- "G-19-20"
fossil_scores_info_8pc[fossil_scores_info_8pc$plotting %in% c("G-10-11", "G-11"),]$plotting <- "G-10-11"
fossil_scores_info_8pc[fossil_scores_info_8pc$plotting %in% c("G-8-9", "G-9"),]$plotting <- "G-8-9"

# chi-squared
subset_8pc <- fossil_scores_info_8pc %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) 
subset_8pc$plotting <- factor(subset_8pc$plotting, levels = c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29"))
subset_8pc <- table(subset_8pc$plotting, subset_8pc$likely_category)
set.seed(1212)
chisq.test(subset_8pc, simulate.p.value = T)

# plot
submember_plot_clean_8pc <- fossil_scores_info_8pc %>% filter(plotting %in% c("G-3", "G-4", "G-5", "G-7", "G-8", "G-8-9", "G-10-11", "G-12", "G-12-13", "G-13", "G-19-20", "G-27-28", "G-28-29")) %>% group_by(plotting, likely_category) %>% tally() %>%
  ggplot()+
  geom_col(aes(x = n, y = plotting, fill = likely_category), position = "fill")+
  scale_fill_brewer(palette = "BrBG", direction = -1, name = "Predicted category")+
  theme_classic(base_size = 16)+
  xlab("Proportion of specimens")+
  ylab("Submember")+
  annotate(geom = "text", x = 1.03, y = as.vector(1:(length(subset_samples))), label = subset_samples, hjust = 0.2, size = 4)+
  theme(legend.position = "bottom")

submember_plot_clean_8pc
```

Combine 30PC and 8PC CVA plots:
```{r fig.height=9, fig.width=25}
(submember_plot_clean_30pc | submember_plot_clean_8pc) + plot_layout(axis_titles = "collect", axes = "collect_y", guides = "collect") & theme(legend.position = "bottom") 
```
